---
title: "NACC preprocessing"
author: "Yihan Wang"
date: "2025-03-17"
output: word_document
---
```{r}
# rm(list = ls())
# setwd("/Users/yihanwang/Documents/Florey/projects/NACC-CAA/code")
```
```{r}

library(dplyr)
library(ggplot2)
library(readxl)
library(zoo)
```


```{r}
nacc_df_raw<- read.csv("../raw_data/investigator_ftldlbd_nacc66.csv", check.names = FALSE)

```

Step 1: Understanding the dataset
- dimension
```{r}
dim(nacc_df_raw)
```

- how many participants
```{r}
length(unique(nacc_df_raw$NACCID))
```

```{r}
head(nacc_df_raw$NACCID)
```

For further sorting or analysis, convert the IDs to the numeric only
```{r}
nacc_df_raw$NACCID <- as.integer(gsub("NACC", "", nacc_df_raw$NACCID))
head(nacc_df_raw$NACCID)
```


```{r}
replacements <- c(-4, -4.4, 8, 9, 88, 99, 97, 98, 96, 95, 
                  888, 999, 8888, 9999, 88.8, 888.8, 99.9, 777)

nacc_df <- nacc_df_raw %>% 
  select(NACCID, where(is.numeric))%>%
  mutate(across(where(is.numeric), ~ replace(. , . %in% replacements, NA)))

```

get average age 
```{r}
baseline_data <- nacc_df %>%
  group_by(NACCID) %>%
  arrange(NACCAGE) %>%
  slice(1) %>% 
  ungroup()

print(mean(baseline_data$NACCAGE, na.rm = TRUE))

```

Some visualizations
```{r}
nacc_df_clean <- subset(nacc_df, !is.na(NACCAGE))

ggplot(nacc_df_clean, aes(x = NACCAGE)) +
  geom_histogram(aes(y = ..density..), binwidth = 2, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_density(color = "red", linewidth = 1.2) +
  labs(title = "Histogram and KDE of NACCAGE", x = "NACCAGE", y = "Density")

```
```{r}
nacc_df_clean <- subset(baseline_data, !is.na(NACCAGE))
ggplot(nacc_df_clean, aes(x = NACCAGE)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "lightblue", color = "black", alpha = 0.8) +
  geom_density(color = "red", linewidth = 1.2) +
  labs(title = "Histogram and KDE of NACCAGE", x = "NACCAGE", y = "Density")

```


```{r}
nacc_df_clean <- subset(nacc_df, !is.na(NACCAGE) & !is.na(CDRSUM))

# Now plot the trajectories without missing values
ggplot(nacc_df_clean, aes(x = NACCAGE, y = CDRSUM, group = NACCID)) +
  geom_line(alpha = 0.5) +
  labs(title = "Age Trajectories for Each Participant",
       x = "Age (NACCAGE)",
       y = "CDR-SOB") +
  theme_minimal()

```
```{r}

```

Get number of visits
```{r}
visit_counts <- nacc_df %>%
  group_by(NACCID) %>%
  summarise(visit_count = n())

# Create a bar chart showing the frequency of participants by their number of visits
ggplot(visit_counts, aes(x = as.factor(visit_count))) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Visit Counts per Participant",
       x = "Number of Visits",
       y = "Number of Participants") +
  theme_minimal()
```
Step 2: know for the targets and select the features
```{r}
unique_ids <- unique(nacc_df[c("NACCID", "NPADNC")])
table(unique_ids$NPADNC)

```


```{r}
feature_set <- read_excel("../raw_data/NACC_CAA_path_variable.xlsx", sheet = "python")$Feature
demo_set <- read_excel("../raw_data/NACC_CAA_path_variable.xlsx", sheet = "demographic")$Feature
pathology_set <- read_excel("../raw_data/NACC_CAA_path_variable.xlsx", sheet = "pathology")$Feature
comorbidity_set <- read_excel("../raw_data/NACC_CAA_path_variable.xlsx", sheet = "comorbidity")$Feature
medication_set <- read_excel("../raw_data/NACC_CAA_path_variable.xlsx", sheet = "medication")$Feature
```

select those who have died
```{r}
NACC_sorted <- nacc_df %>% arrange(NACCID, NACCAGE)
participants_with_death <- unique(NACC_sorted$NACCID[NACC_sorted$NACCDIED == 1])
NACC_filter <- NACC_sorted %>% filter(NACCID %in% participants_with_death)
NACC_filter <- NACC_filter[, feature_set]
cat("There are", length(unique(NACC_filter$NACCID)), "unique participant died.\n")
print(dim(NACC_filter))
```

Step 3: imputations - logic/based on knowledge
```{r}
fill_na <- function(x) {
  if (all(is.na(x))) {
    return(x)
  } else {
    x <- na.locf(x, na.rm = FALSE) # Forward fill: replace NA with the last observed non-NA value
    x <- na.locf(x, fromLast = TRUE, na.rm = FALSE) # Backward fill: replace remaining NA with the next non-NA value
    return(x)
  }
}

NACC_filter <- NACC_filter %>%
  group_by(NACCID) %>%
  mutate(across(all_of(demo_set), fill_na)) %>%
  ungroup()
```

```{r}
custom_fill <- function(x) {
  if (any(x == 1, na.rm = TRUE)) {
    return(1)
  } else if (any(!is.na(x))) {
    return(0)
  } else {
    return(NA)
  }
}
```

```{r}
NACC_filter <- NACC_filter %>%
  group_by(NACCID) %>%
  mutate(across(all_of(comorbidity_set), ~ custom_fill(.))) %>%
  mutate(across(all_of(medication_set), ~ custom_fill(.))) %>%
  ungroup()
```

select only the last record for each participants
```{r}
NACC_filter_sorted <- NACC_filter %>%
  arrange(NACCID, NACCAGE)

NACC_filter_cross <- NACC_filter_sorted %>%
  group_by(NACCID) %>%
  slice({
    if (is.na(first(NACCDAGE))) {
      n()
    } else {
      which.min(abs(NACCAGE - first(NACCDAGE)))
    }
  }) %>%
  ungroup()
```




##############################
```{r}
library(dplyr)

# Define the function to calculate NPTDPSUM for one row.
calculate_NPTDPSUM <- function(row) {
  cols <- c("NPTDPA", "NPTDPB", "NPTDPC", "NPTDPD", "NPTDPE")
  
  # Check if all specified columns are NA; if yes, return NA.
  if (all(is.na(row[cols]))) {
    return(NA_real_)
  }
  
  # Replace NA with 0 and check if the sum is 0; if yes, return 0.
  if (sum(ifelse(is.na(row[cols]), 0, row[cols])) == 0) {
    return(0)
  }
  
  # Create a mapping from column names to numeric values.
  mapping <- c(NPTDPA = 1, NPTDPB = 2, NPTDPC = 3, NPTDPD = 4, NPTDPE = 5)
  max_value <- 0
  
  # Loop through each column in the mapping.
  for (col in names(mapping)) {
    # If the column is not NA and equals 1, update max_value.
    if (!is.na(row[[col]]) && row[[col]] == 1) {
      max_value <- max(max_value, mapping[[col]])
    }
  }
  
  # Return max_value if greater than 0, else NA.
  if (max_value > 0) {
    return(max_value)
  } else {
    return(NA_real_)
  }
}

# Apply the function row by row.
NACC_filter_cross <- NACC_filter_cross %>%
  rowwise() %>%
  mutate(
    # Convert selected columns to a named list before passing to the function.
    NPTDPSUM = calculate_NPTDPSUM(as.list(c_across(all_of(c("NPTDPA", "NPTDPB", "NPTDPC", "NPTDPD", "NPTDPE")))))
  ) %>%
  ungroup() %>%
  # Remove the original columns
  select(-NPTDPA, -NPTDPB, -NPTDPC, -NPTDPD, -NPTDPE)

```


```{r}
NPOFTD_cols <- c("NPOFTD", "NPOFTD1", "NPOFTD2", "NPOFTD3", "NPOFTD4", "NPOFTD5")

# Create a new column 'NPOFTDSUM' based on the following rules:
# - If any value in the row (among NPOFTD_cols) is 1, then NPOFTDSUM is 1.
# - If every value is 0 (with no missing values), then NPOFTDSUM is 0.
# - Otherwise, NPOFTDSUM is set to NA.
NACC_filter_cross <- NACC_filter_cross %>%
  rowwise() %>%
  mutate(NPOFTDSUM = {
    # Get the current row values for the specified columns
    values <- c_across(all_of(NPOFTD_cols))
    if (any(values == 1, na.rm = TRUE)) {
      1
    } else if (all(!is.na(values)) && all(values == 0)) {
      0
    } else {
      NA_real_
    }
  }) %>%
  ungroup()

# Define the column names for the second sum
FTD_sum_cols <- c("NPOFTDSUM", "NPFTDTAU", "NPFTDTDP")

# Create a new column 'NPFTDSUM' based on the following rules:
# - If any value in the row (among FTD_sum_cols) is 1, then NPFTDSUM is 1.
# - If every value is 0 (with no missing values), then NPFTDSUM is 0.
# - Otherwise, NPFTDSUM is set to NA.
NACC_filter_cross <- NACC_filter_cross %>%
  rowwise() %>%
  mutate(NPFTDSUM = {
    # Get the current row values for the specified columns
    values <- c_across(all_of(FTD_sum_cols))
    if (any(values == 1, na.rm = TRUE)) {
      1
    } else if (all(!is.na(values)) && all(values == 0)) {
      0
    } else {
      NA_real_
    }
  }) %>%
  ungroup()


```


```{r}
library(dplyr)

NACC_with_CAA_cross <- NACC_filter_cross %>%
  group_by(NACCID) %>%
  mutate(CAA_death = last(NACCAMY)) %>%
  slice(1) %>%
  ungroup() %>%
  filter(!is.na(CAA_death))


```


```{r}
missing_perc <- sort(colMeans(is.na(NACC_with_CAA_cross)) * 100, decreasing = TRUE)
missing_perc


```

```{r}
library(dplyr)

# 1. Drop rows with missing values in the specified columns
NACC_with_CAA_cross <- NACC_with_CAA_cross %>%
  filter(!is.na(NPADNC),
         !is.na(NACCLEWY),
         !is.na(CAA_death),
         !is.na(NPTDPSUM),
         !is.na(NPTHAL),
         !is.na(NACCBRAA),
         !is.na(NACCNEUR)) %>%
  filter(!is.na(NACCAGE),
         !is.na(NACCDAGE),
         !is.na(NACCAGEB))

# 2. Print the number of unique participants based on NACCID
cat("There are", n_distinct(NACC_with_CAA_cross$NACCID), "unique participants died with CAA value.\n")

# 3. Remove rows where the proportion of missing values is 20% or higher
NACC_with_CAA_cross <- NACC_with_CAA_cross %>%
  filter(rowMeans(is.na(.)) < 0.2)

# 4. Identify columns with less than 20% missing values
final_cols <- names(which(colMeans(is.na(NACC_with_CAA_cross)) < 0.2))

# 5. Keep only these columns in the data frame
NACC_with_CAA_cross <- NACC_with_CAA_cross[, final_cols]

# 6. Print the updated number of unique participants and the dimensions of the data frame
cat("There are", n_distinct(NACC_with_CAA_cross$NACCID), "unique participants died with CAA value.\n")
cat("Data frame dimensions:", dim(NACC_with_CAA_cross), "\n")


```




```{r}
last_records <- nacc_df %>%
  group_by(NACCID) %>%
  slice_tail(n = 1) %>%
  ungroup()

# Create a cross-tabulation between CAA_death and NACCUDSD
crosstab <- table(last_records$CAA_death, last_records$NACCUDSD)
print(crosstab)

```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

